An algorithm for the optimization of finite element integration loops

### 2016-11-30 TOMS_Paper/assembly.tex
Slightly smaller plots
Add stuff about flops performed and data transfer

Following the methodology adopted in~\citeN{quadrature1}, we vary the following parameters:
\begin{itemize}
\item the polynomial degree of test, trial, and coefficient (or ``pre-multiplying'') functions, $q \in \lbrace1, 2, 3, 4\rbrace$
\item the number of coefficient functions $\mathrm{nf} \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
While constants of our study are
\begin{itemize}
\item the space of test, trial, and coefficient functions: Lagrange
\item the mesh: tetrahedral with a total of 4374 elements
\item exact numerical quadrature (we employ the same scheme used in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule)
\end{itemize}

@@

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
\caption{Performance evaluation for the \textit{mass} matrix. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a \textit{Helmholtz} equation. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
\end{figure}

\begin{figure}
\includegraphics[scale=0.77]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in an \textit{elastic} model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a \textit{hyperelastic} model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
\end{figure}

@@

\paragraph{Mass matrix}
We start with the simplest of the bilinear forms investigated, the mass matrix. Results are in Figure~\ref{fig:mass}. We first notice that the lack of improvements when $q=1$ is due to the fact that matrix insertion outweighs local assembly. For $q \geq 2$, \texttt{cfO2} generally shows the highest speed-ups. It is worth noting why \texttt{auto} does not always select the fastest implementation: \texttt{auto} always opts for \texttt{tens}, while for $\mathrm{nf} \geq 2$ \texttt{quad} tends to be preferable. On the other hand, \texttt{cfO2} always makes the optimal decision about whether to apply pre-evaluation or not. Surprisingly, despite the simplicity of the form, the performance of the various code generation systems can differ significantly.

\paragraph{Helmholtz}
As in the case of Mass matrix, when $q=1$ the matrix insertion phase is dominant. For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors. In particular:
\begin{description}
\item[$\mathrm{nf}=0$] pre-evaluation makes \texttt{cfO2} notably faster than \texttt{cfO1}, especially for high values of $q$; \texttt{auto} correctly selects \texttt{tens}, which is comparable to \texttt{cfO2}. 
\item[$\mathrm{nf}=1$] \texttt{auto} picks \texttt{tens}; the choice is however sub-optimal when $q=3$ and $q=4$. This can indirectly be inferred from the large gap between \texttt{cfO2} and \texttt{tens/auto}: \texttt{cfO2} applies sharing elimination, but it correctly avoids pre-evaluation because of the excessive expansion cost.
\item[$\mathrm{nf}=2$ and $\mathrm{nf}=3$] \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens}); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
\end{description}

The slow-downs (or marginal improvements) seen in a small number of cases exhibited by \texttt{ufls} can be attributed to the presence of sharing in the generated code.

@@

\paragraph{Elasticity}
The results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The main observation is that \texttt{cfO2} never triggers pre-evaluation, although in some occasions it should. To clarify this, consider the test case $\langle \mathrm{nf}=0,\ q=2 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}. \texttt{cfO2} finds pre-evaluation profitable in terms of operation count, although it is eventually not applied to avoid exceeding $T_H$. However, running the same experiments with $T_H = \operatorname{size}(\mathrm{L3})$ resulted in a dramatic improvement, even higher than that obtained by \texttt{tens}. The reason is that, despite exceeding $T_H$ by roughly 40$\%$, the saving in operation count is so large (5$\times$ in this specific problem) that pre-evaluation would in practice be the winning choice. This suggests that our objective function should be improved to handle the cases in which there is a significant gap between potential cache misses and reduction in operation count.

@@

>>>>

Following the methodology adopted in~\citeN{quadrature1}, we vary the following parameters:
\begin{itemize}
\item the polynomial degree of test, trial, and coefficient (or ``pre-multiplying'') functions, $q \in \lbrace1, 2, 3, 4\rbrace$
\item the number of coefficient functions $\mathrm{nf} \in \lbrace0, 1, 2, 3\rbrace$
\end{itemize}
While constants of our study are
\begin{itemize}
\item the space of test, trial, and coefficient functions: Lagrange
\item the mesh: tetrahedral with a total of 4374 elements
\item exact numerical quadrature (we employ the same scheme used in~\citeN{quadrature1}, based on the Gauss-Legendre-Jacobi rule)
\item double-precision arithmetic
\end{itemize}

@@

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.70]{perf-results/mass}}
\caption{Performance evaluation for the \textit{mass} matrix. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.70]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a \textit{Helmholtz} equation. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
\end{figure}

\begin{figure}
\includegraphics[scale=0.70]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in an \textit{elastic} model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.70]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a \textit{hyperelastic} model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
\end{figure}

@@

\paragraph{Mass matrix}
We start with the simplest of the bilinear forms investigated, the mass matrix. Results are in Figure~\ref{fig:mass}.

We first notice the lack of improvements when $q=1$. The operation count of \texttt{cfO2} and \texttt{tensor} is roughly 9$\times$ lower than that of the alternatives (tens instead of hundreads operations), but nontheless no gain in execution time is obtained. This is due to the fact that local assembly is memory-bound. For instance, in the simplest case $nf=0$, in addition to the determinant of the Jacobian inverse, \texttt{tens} only requries three unique multiplications to evaluate the local element matrix (26 operations in total), whereas almost a hundred memory accesses are performed. In particular, we have: 12 loads for the vertex coordinates; for each vertex, 1 load to indirectly access its data; for each of the 16 local elements, 4 loads (optimistically) and 1 store to update an entry in the global matrix in compressed sparse row format. This unsurprisingly leads to a very low arithmetic intensity (i.e., $\frac{\text{operation count}}{\text{bytes moved}}$), which collocates the kernels in a memory-bound regime.

For $q \geq 2$, instead, the transformation systems impact the performance, and \texttt{cfO2} generally shows the highest speed-ups. It is worth noting that \texttt{auto} does not always select the fastest implementation, as it always opts for \texttt{tens}, while for $\mathrm{nf} \geq 2$ \texttt{quad} tends to be preferable. On the other hand, \texttt{cfO2} always makes the optimal decision about whether to apply pre-evaluation or not. Unexpectedly, in spite of the simplicity of the form, the performance of the various code generation systems can differ significantly.

\paragraph{Helmholtz}
As in the case of mass matrix, when $q=1$ the matrix insertion phase is dominant. For $q \geq 2$, the general trend is that \texttt{cfO2} outperforms the competitors. In particular:

\begin{description}
\item[$\mathrm{nf}=0$] pre-evaluation makes \texttt{cfO2} notably faster than \texttt{cfO1}, especially for high values of $q$; \texttt{auto} correctly selects \texttt{tens}, which is comparable to \texttt{cfO2}. 
\item[$\mathrm{nf}=1$] \texttt{auto} picks \texttt{tens}; the choice is however sub-optimal when $q=3$ and $q=4$. This can indirectly be inferred from the large gap between \texttt{cfO2} and \texttt{tens/auto}: \texttt{cfO2} applies sharing elimination, but it correctly avoids pre-evaluation because of the excessive expansion cost.
\item[$\mathrm{nf}=2$ and $\mathrm{nf}=3$] \texttt{auto} reverts to \texttt{quad}, which would theoretically be the right choice (the flop count is much lower than in \texttt{tens}); however, the generated code suffers from the presence of indirection arrays, which break autovectorization and ``traditional'' code motion.
\end{description}

In the case $\langle \mathrm{nf}=0,\ q=2 \rangle$, the 2$\times$ speed-ups achieved by \texttt{cfO2} and \texttt{tens} are due to the remarkable drop in operation count achieved by pre-evaluating reductions: from order $10^4$ (\texttt{ufls} $3 \cdot 10^4$, \texttt{cfO1} $2 \cdot 10^4$) to order $10^3$. By unrolling loops and eliminating operations over zeros, \texttt{tens} eventually executes fewer operations than \texttt{cfO2} (roughly, $10^3$ instead of $2 \cdot 10^3$), but the performance is basically identical because of matrix insertion, which plays again the limiting role. With $10^2$ entries to be updated in the global matrix, following the same procedure of the previous paragraph, it is straightforward to verify that $\frac{\text{operation count}}{\text{bytes moved}} \ll 1$, a typical symptom of memory-boundedness.

Another relevant problem instance is $\langle \mathrm{nf}=0,\ q=4 \rangle$, in which \texttt{cfO2} performs 1.5$\times$ more operations than \texttt{tens}, but still exhibits a better execution time. This is again caused by the chosen code generation strategies: while \texttt{tens} prefers to fully unroll the computation to detect all avoidable operations, \texttt{cfO2} preserves the loop structure. This choice mitigates the memory transfer (especially between L1 cache and registers) by maximizing register locality.

The slow-downs (or marginal improvements) seen in a small number of cases exhibited by \texttt{ufls} can be attributed to the presence of sharing in the generated code. Further, the factorization applied by \texttt{ufls} hides code motion opportunities that the underlying compiler would otherwise have been able to exploit.

@@

\paragraph{Elasticity}
The results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The main observation is that \texttt{cfO2} never triggers pre-evaluation, although in some occasions it should. To clarify this, consider the test case $\langle \mathrm{nf}=0,\ q=4 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}. \texttt{cfO2} finds pre-evaluation profitable in terms of operation count, although it is eventually not applied to avoid exceeding $T_H$. However, running the same experiments with $T_H = \operatorname{size}(\mathrm{L3})$ resulted in a dramatic improvement, even higher than that obtained by \texttt{tens}. The reason is that, despite exceeding $T_H$ by roughly 40$\%$, the saving in operation count is so large (5$\times$ in this specific problem) that pre-evaluation would in practice be the winning choice. This suggests that our objective function should be improved to handle the cases in which there is a significant gap between potential cache misses and reduction in operation count.

@@

We also tried to relate the achieved performance to the operation counts obtained by each transformation system. We here limit ourselves to consider the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups are generally massive. For the untransformed, original code, at the source level $37 \cdot 10^9$ operations are counted. However, this number is in practice much lower, since by inspection of the assembly code generated by the underlying compiler, it is clear that common sub-expressions elimination and loop-invariant code motion have aggressively been applied. The actual operation count, however, is definitely larger than the $10^8$ operations performed by \texttt{ufls}. \texttt{cfO2} brings the operation count down to $3 \cdot 10^7$; that is, more than 3$\times$ lower than \texttt{ufls}. This is completely in line with the drop in execution times, with \texttt{cfO2} outperforming \texttt{ufls} by more than 3$\times$.

Unsurprisingly, the memory transfer impact is less relevant in this hyperelastic than in all previous forms. With $\langle \mathrm{nf}=0,\ q=1 \rangle$, \texttt{cfO2} is already 1.4$\times$ faster than the untransformed code, and the gain increases with both $nf$ and $q$.
